### 2022-06-22 paper.Rmd
remove unused footnote
some minor editing

[^averagesum]:

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the elements. A different anticlusting approach is based on clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known clustering method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. The k-plus objective is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Hence, the k-plus criterion is not a single criterion for anticlustering, but a familiy of objectives implementing different requirements with regard to between-group similarity. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two pratical are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

In anticlustering applications, we impose an additional constraint on the group sizes $n_k$, where the most common restriction requires that all groups have equal size:

@@

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective, i.e., when just appending the squared-distance-features to the data set. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ have much larger values, on average, than the raw values $x_{ij}$. Therefore, the $\mathit{SSE_{Var}}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

Thus, to maximize between-groups similarity with regard to covariance structure, it is possible to add an additional variable for each pair of variables $x$ and $y$, computed as $(x_i - \overline{\mathbf{x}}) (y_i -\overline{\mathbf{y}})$. This requires computing $\binom{N}{2}$ new variables and thus turns out be be rather costly: The number of additional variables needed to minimize differences with regard to covariance structure grows quadratically with increasing number of variables; the number of additional variables needed to minimize differences with regard to distribution moments only grows linearly with increasing number of variables.

@@

To thorougly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to the mean, variance and covariance structure. To implement all k-plus objectives, the simple unweighted sum outlined above approach was employed. Thus, when optimizing a k-plus criterion, the original data matrix was extended to incorporate the additional criteria. Subsequently, the original k-means criterion $SSE$ was maximized on the basis of the augmented data set.

@@

```{r Figure1, fig.width = 8, fig.height = 10, fig.cap = "Simulation results by $N$."}

# Plot the results, by N
ldf %>% 
  group_by(method, Objective, N) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>%
  ggplot(aes(x = N, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16)

@@

Table 1 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means, but disregards all other distribution characteristics. K-plus anticlustering also addresses $\Delta \mathit{M}$ quite well and at the same time tends to minimize discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$). With regard to minimizing $\Delta \mathit{SD}$, all three k-plus approaches outperform all other methods. As Figure 1 shows, with increasing $N$, k-means and all k-plus methods converge on minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time.

While k-plus anticlustering is capable of addressing multiple objectives simultaneously, it is apparent that increasing the number of optimization criteria aggravates fulfilling each single criterion. K-plus skew/kurtosis and k-plus correlation are worse at minimizing discrepancy with regard to standard deviations than standard k-plus, which is slightly worse at minimizing $\Delta \mathit{M}$ than k-means anticlustering. Still, every k-plus method does what it is supposed to do: k-plus skew/kurtosis is best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation is best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintain a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to be much better than diversity anticlustering with regard to skew and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis is even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperforms diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitates to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering is outperformed. As as side note, apart from k-plus correlation, diversity anticlustering is the only method that equalizes correlation structure among groups. As far as I know, this observation has not been made previously, even though the maximum diversity problem has been widely studied.

>>>>

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the data set. A different anticlusting approach is based on clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. The k-plus criterion is not a single criterion for anticlustering, but a familiy of objectives implementing different requirements with regard to between-group similarity. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two pratical are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

In anticlustering applications, an additional constraint is usually imposed on the group sizes $n_k$, where the most common restriction requires that all groups have equal size:

@@

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective, i.e., when just appending the squared-distance-features to the data set. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ usually have much larger values, on average, than the raw values $x_{ij}$. Therefore, the $\mathit{SSE_{Var}}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

Thus, to maximize between-group similarity with regard to covariance structure, it is possible to add an additional variable for each pair of variables $x$ and $y$, computed as $(x_i - \overline{\mathbf{x}}) (y_i -\overline{\mathbf{y}})$. This requires computing $\binom{N}{2}$ new variables and thus turns out be be rather costly: The number of additional variables needed to minimize differences with regard to covariance structure grows quadratically with increasing number of variables; the number of additional variables needed to minimize differences with regard to distribution moments only grows linearly with increasing number of variables.

@@

To thoroughly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to the mean, variance and covariance structure. To implement all k-plus objectives, the simple unweighted sum outlined above approach was employed. Thus, when optimizing a k-plus criterion, the original data matrix was extended to incorporate the additional criteria. Subsequently, the original k-means criterion $SSE$ was maximized on the basis of the augmented data set.

@@

```{r Figure1, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by $N$."}

# Plot the results, by N
ldf <- ldf %>% 
  group_by(method, Objective, N) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  )

ldf %>%
  ggplot(aes(x = N, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16)

@@

Table 1 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but disregarded all other distribution characteristics, where it did not outperform a random assignment. K-plus anticlustering also addressed $\Delta \mathit{M}$ quite well and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); all three k-plus approaches outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

As Figure 1 shows, with increasing $N$, k-means and all k-plus methods converge on minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, it is apparent that increasing the number of optimization criteria aggravated fulfilling each single criterion. K-plus skew/kurtosis and k-plus correlation performed worse at minimizing discrepancy with regard to standard deviations than standard k-plus, which itself performed slightly worse than k-means anticlustering. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitates to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat addressed all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. As as side note, apart from k-plus correlation, diversity anticlustering was the only method that equalizes correlation structure among groups. As far as I know, this observation has not been made previously, even though the maximum diversity problem has been widely studied.
